Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-07 paper.tex
Catch bug in simulated blocking. Add performance comparison section.

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{subcaption}
\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}

@@

Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces a negligible latency overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.

@@

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

  \item \textbf{Initialization:}
        \begin{itemize}
          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
          \item The first thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
        \end{itemize}

@@

\john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance. For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance. Perhaps worth pointing out. Does the flag-unset have any ordering constraints in terms of correctness?}
\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
\thomas{No furher revisions required/OK to remove comments?}.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size of 256 threads, where each thread processes 4 vectors of size 4, resulting in 16 elements per thread and a total work tile size of 4096. We find a maximum spin count of 4 as offering good responsiveness to blocking, while minimizing false-positives fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the minimum subgroup size specified by the WebGPU standard, which is 4. Despite the additional shared memory required for conflict avoidance and fallback operations, total shared memory usage remains modest at only 1 KB. For all tests, we use an input size of $2^{25}$.

Something about devices here.
\subsection{Performance Comparison}
To evaluate the performance of \emph{Decoupled-Fallback}, we compare it against two key benchmarks: \emph{Reduce-Then-Scan} (RTS) and a \emph{Memcopy} operation. RTS serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique. Memcopy is included as a point of reference, as it shares the same $2n$ global memory movement required by scan operations.

\thomas{Forgot about tables. Mega table incoming.}

\thomas{Performance is good}

@@

\subsection{Emulated Blocking}
To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions. Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
\begin{figure}
  \centering
  \begin{subfigure}{.9\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated}
    \label{fig:fallbacks_initiated}
  \end{subfigure}
  \begin{subfigure}{.9\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins}
    \label{fig:total_spins}
  \end{subfigure}
  \begin{subfigure}{.9\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
    \caption{Lookback Length}
    \label{fig:lookback_length}
  \end{subfigure}
  \begin{subfigure}{.9\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
    \caption{Speed vs Memcopy}
    \label{fig:execution_time}
  \end{subfigure}
  \caption{Simulated Blocking\john{Text in these plots---axis labels, etc.---needs to be larger. If it's significantly smaller than body text, which it is, it's just not legible.}}
  \label{fig:vertical_images}
\end{figure}

In Figure~\ref{fig:fallbacks_initiated}, we measure the number of fallbacks proportional to the number of workgroups launched, and verify that we are indeed inducing fallbacks. In Figure~\ref{fig:total_spins} we measure \thomas{this should be spins per workgroup it's a better metric}. In Figure~\ref{fig:lookbackLength} we see that. Finally, in Figure~\ref{fig:execution_time}, we can see that, even when \emph{half} the workgroups are deadlocking, \emph{Decoupled-Fallback} still achieves speeds of . . .

@@

\clearpage
\appendix
\section{Artifact}

\subsection{API Choice}
As rust is significantly more easy to build across devices we chose the WGPU api.

\subsection{Device Choices}
We chose a diverse range of devices to illustrate

\subsection{Baseline Comparisons}
Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes. Similarly, as there are no existing WebGPU implementations of RTS, we rely on our own optimized version for comparison. To ensure a competitive evaluation, our RTS implementation employs a variable-launch main kernel with a single workgroup performing a serial spine scan, as we found raking to be less performant in practice. Apart from this adjustment, the intraworkgroup implementation in RTS is identical to that used in \emph{Decoupled-Fallback}.

\subsection{Availability}
We provide our artifact in the following GitHub repository: 
\thomas{Will anonymize with GithubAnonymous once we are ready.} 
We provide both Dawn and WGPU implementations for testing. However, we recommend Dawn, and only use WGPU for ease of native testing on Android.

@@

\subsubsection{Device Requirements}
Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory.

>>>>

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{makecell}
\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}

@@

Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.

@@

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

  \item \textbf{Initialization:}
        \begin{itemize}
          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
          \item The first thread initializes a \emph{previous reduction} variable in registers with $\varepsilon$ to hold the accumulated reductions of preceding tiles.
          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
        \end{itemize}

@@

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, we find a maximum spin count of 4 to effectively balance responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU.}
\begin{table}[t]
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & FPG & Occupancy & Subgroup Size \\  
    \midrule
    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8 - 32  \\  
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\  
    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\  
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\  
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32 - 64  \\  
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}). \emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique. \emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound.

\begin{table*}
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M1 Pro Max & Nvidia 2080 Super & AMD 7900 XT \\  
    \midrule
    H-Mean \emph{Memcopy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
    H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\  
    H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\  
    \emph{CSDLDF}/\emph{Memcopy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\  
    \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\  
    Avg Spins per Wg    & 0.306    & 1.380    & 0.418    & 0.418    & 0.133    \\  
    Avg Lookback Length per Wg  & 1.725    & 3.510    & 13.139   & 13.139   & 20.867   \\  
    Avg Total Fallbacks Initiated      & 260.364   & 2202.045  & 88.384   & 88.384   & 10.127   \\  
    Avg Total Successful Insertions    & 6.084     & 57.875    & 1.205    & 1.205    & 1.001    \\
    \bottomrule
  \end{tabular*}
  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
  \label{tab:results}
\end{table*}

As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices. Since \emph{Memcopy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput. This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} slightly outperforms \emph{Memcopy}, suggesting improved memory structuring due to compute operations. On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%. However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory bandwidth bound, limiting further gains.

@@

\subsection{Simulated Blocking}
To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions. Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated}
    \label{fig:fallbacks_initiated}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins}
    \label{fig:total_spins}
  \end{subfigure}

  \vspace{0.2em}

  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
    \caption{Lookback Length}
    \label{fig:lookback_length}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
    \caption{Throughput vs Baseline}
    \label{fig:execution_time}
  \end{subfigure}

  \caption{Simulated Blocking}
  \label{fig:vertical_images}
\end{figure*}

In Figure~\ref{fig:fallbacks_initiated}, we measure the number of fallbacks proportional to the number of workgroups launched, and verify that we are indeed inducing fallbacks. In Figure~\ref{fig:total_spins} we measure \thomas{this should be spins per workgroup it's a better metric}. In Figure~\ref{fig:lookback_length} we see that. Finally, in Figure~\ref{fig:execution_time}, we can see that, even when \emph{half} the workgroups are deadlocking, \emph{Decoupled-Fallback} still achieves speeds of . . .

@@

\appendix
\section{Artifact}
\subsection{Baseline Comparisons: \emph{Memcopy}}
\label{sec:memcopy}
\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{l r r}
    \toprule
    Device & \makecell{Memory \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\  
    \midrule
    Intel HD620 (Single-/Dual-Channel) & 17.1 / 34.1  & 2.13e9 / 4.26e9 \\  
    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2  & 6.40e9 \\  
    Apple M1 Pro Max   & 200  & 25.0e9 \\  
    Nvidia 2080 Super  & 448  & 56.0e9 \\  
    AMD 7900 XT        & 800  & 100.0e9 \\  
    \bottomrule
  \end{tabular}
  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
  \label{tab:memory_bandwidth}
\end{table}
Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcopy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation. It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above. We make a number of observations:
\begin{itemize}
  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec. Thus, our observed bandwidth of 1.498e9 is reasonable.
  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec. The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging. Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
\end{itemize}

\subsection{Baseline Comparisons: \emph{Reduce then Scan}}
\label{sec:rts}
As we desire the most competitve baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{CSDLDF}. Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing. However, empirical testing revealed that raking was consistently outperformed—by up to $\sim$15\% on lower-end devices—by a simpler approach that processes the spine serially with a single workgroup. Furthermore, our \emph{RTS} performance closely aligns with the theoretically expected $O(3n)$ global memory movement.

\subsection{Availability}
We provide our artifact in the following GitHub repository: 
\thomas{Will anonymize with GithubAnonymous once we are ready.}

@@

\subsubsection{Device Requirements}
Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory. Requires at least $\sim$2GB of disk memory to build Dawn.
